A Bayesian search to find \\high-mass black holes in LIGO data

### 2020-08-20 main.tex
Updates from Overleaf

\documentclass[%
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4}

@@

\newcommand{\prior}{{\sc prior}\xspace}
\newcommand{\bilby}{{\sc Bilby}\xspace}
\newcommand{\bilbypipe}{{\sc bilby\_pipe}\xspace}
\newcommand{\bilbypipegdb}{\texttt{bilby\_pipe\_gracedb}\xspace}
\newcommand{\pbilby}{{\sc pBilby}\xspace}
\newcommand{\dynesty}{{\sc dynesty}\xspace}
\newcommand{\cpnest}{{\sc cpnest}\xspace}
\newcommand{\ptemcee}{{\sc ptemcee}\xspace}
\newcommand{\gwpy}{{\sc GWpy}\xspace}
\newcommand{\imrphenomp}{{\sc IMRPhenomPv2}\xspace}
\newcommand{\pycbc}{{\sc PyCBC}\xspace}
\newcommand{\msun}{{M${}_\odot$}\xspace}

@@

\begin{abstract}
The detection of high mass black holes ( $>100$ \msun) will shed light on the formation of supermassive black holes and thus galaxy formation. Although LIGO is sensitive to the merger of binary black holes with total masses up to 500 \msun, the largest total mass detected so far is approximately $80$ \msun. A possible explanation for the absence of high mass events may be due to their misclassification as short-duration instrumental noise transients. Short-duration instrumental transients mimic the short-duration gravitational-wave signals from high-mass binary black hole mergers. Here we demonstrate that a new search method utilising Bayesian inference could be a more sensitive tool for detecting high-mass binary black hole mergers as compared to traditional match-filtering. We have applied this technique on the high-mass triggers during LIGO's second observing run to investigate the possibility of discovering new gravitational-wave signals from an entirely new class of high mass black hole binaries.

@@

Since the 1970s, there has been an accumulation of evidence for stellar-mass and supermassive black holes. In 2019, the Event Horizon Telescope provided the first visual evidence of the supermassive black hole M87  \cite{m87photo}.  As of January 2020, LIGO, Virgo and IAS have confirmed more than a dozen binary black hole systems \cite{GWTC1, IAS0, IAS1, IAS2}. These various discoveries have firmly established the existence of stellar-mass black holes, supermassive black holes and binary black hole systems.  Interestingly, there is still no direct evidence for intermediate-mass black holes, the black holes that lie in between stellar-mass and supermassive black hole systems with masses between $10^2-10^6$ \msun. Thus, the existence of intermediate-mass black holes is still speculative. \\

If intermediate-mass black holes are present, the gravitational waves emitted from the merger of a binary black hole system with at least one intermediate-mass black hole (up to a mass of $400$ \msun) will be detectable by the ground-based gravitational-wave network. Gravitational waves from such systems should occur at a rate of $0-10\text{yr}^{-1}$ \cite{fregeau2006imbhbRatePrediction, mandel2008rates,rodriguez2015bbhRatePredictions}.  However, even after conducting a targeted match-filter based search for gravitational waves from intermediate-mass black holes the largest total mass detected so far is approximately $80$ \msun \cite{salemi2019search, abbott2019gwtc}. A possible explanation for the absence of intermediate-mass events may be due to their misclassification as short-duration instrumental noise transients \cite{blip_glitches}. These glitches can mimic astrophysical signals and hence decrease the significance of true gravitational wave events.

One method to account for glitches while searching for gravitational waves from coalescing compact binaries is by utilising a signal consistency test \cite{pycbc_short_duration_transients} or a signal coherence test amongst a network of detectors \cite{bci, bcr_paper,  bcr_gw151216, bayesian_odds}. A signal-coherence test such as the Bayesian coherence ratio (BCR) \cite{bcr_paper} provides an odds ratio comparing the probability that the data contains an astrophysical signal versus the probability that the data contains an incoherent glitch or noise. Although an odds ratio, the BCR can be used  as a ranking statistic to calculate the statistical significance of candidate gravitational-wave signals from compact binary coalescences in the detector data \cite{bcr_paper}.

In this paper we utilise the BCR to calculate $p_\text{astro}$, the probability that a signal is of astrophysical origin [cite],  for all the high-mass (systems with total masses in the range of $55-500$ \msun) candidate gravitational-wave signals from compact binary coalescences in the detector data recorded during aLIGO's second observing run (O2).  We find that (a) our search does not identify any new stellar mass or intermediate mass black holes; (b) high-mass events reported in the GWTC-1, including GW170729 (an event with disputed $p_\text{astro}$ amongst various search pipelines) have high $p_\text{astro}$; and that (c) high-mass events detected from independent groups have low $p_\text{astro}$.

\section{PyCBC}
% PyCBC asks for the 5 citations:
% http://pycbc.org/pycbc/latest/html/credit.html#searches-for-compact-binary-coalescence

 such as \pycbc \cite{pycbc_code} have tried to account for these short-duration glitches \cite{pycbc_short_duration_transients}. 
Search algorithms for gravitational waves from coalescing compact binaries, such as PyCBC \cite{pycbc_code}, have tried to account for these short-duration glitches \cite{pycbc_short_duration_transients}. At its core, PyCBC performs a matched-filter search for binary merger signals using a bank of gravitational-wave template waveforms \cite{pycbc_og0, pycbc_og1, pycbc_og2, pycbc_og3, pycbc_og4, pycbc_og5}.  To account for the short-duration glitches,

 to As demonstrated in \citet{blip_glitches},  High-mass mergers have
very few in-band wave cycles, and hence can easily be mistaken for short-duration instrumental transients.

@@

There are two primary motivations for a targeted high-mass search that utilises Bayesian inference. The first is to help
distinguish gravitataional-waves from high-mass systems from noise-transients, and the second is to try to alleviate the
discord in the significance of some of the ``quiet'' gravitational wave events between various search pipelines.

@@

\caption[Comparing glitches to high-mass candidates.]{Spectrogram of a glitch on the left, and a candidate GW trigger on the right.}\label{fig:compareGlitchAndSignal}
\end{figure}

@@

The goal of this PhD's first project is to enhance the significance of high-mass foreground triggers using a Bayesian
approach and thereby increase the confidence in the detection of gravitational waves from large-mass compact binaries.
If this work does improve the significance of high-mass triggers, we will increase the likelihood of detecting a
high-mass black hole system via gravitational waves. The detection statistic we will utilise is the Bayesian Coherence
Ratio (or BCR), as seen in \citet{bcr_paper}. The BCR is discussed in detail in Appendix\textasciitilde\ref{BayesianCoherenceRatio}. In
short, The \texttt{BCR} is an odds (evidence ratio) comparing two hypotheses: there's a coherent GW in the data, plus Gaussian
noise vs there's only incoherent Gaussian noise or a glitch (modeled as a compact binary coalescence) in each detector
\citep{bcr_paper}.

@@

\caption[p-astro for various O2 foreground triggers]{\label{tab:O2significancesWObcr}p-astro from seveal detection pipelines for a subset of the O2 foreground triggers.}
\centering
\begin{tabular}{llrrrr}
Event & Catalog & PyCBC & GstLAL & cWB & IAS\\
GW170104 & GWTC-1 & 1.00 & 1.00 & 1.00 & 0.99\\
GW170121 & IAS-1 & NA & NA & NA & 0.99\\
GWC170402 & IAS-2 & NA & 0.09 & NA & 0.68\\
GW170403 & IAS-1 & NA & NA & NA & 0.56\\
IMBHC170423 & IMBH-marginal & NA & 0.00 & 1.00 & NA\\
GW170425 & IAS-1 & NA & NA & NA & 0.77\\
GW170729 & GWTC-1 & 0.52 & 0.98 & 0.94 & NA\\
\end{tabular}
\end{table}

@@

\section{Methodology}\label{methodology}
This PhD project intends to provide new methods and insights to studying gravitational waves form high-mass systems.
Thus far, the project has focused on building up the infrastructure to compare the Bayesian Coherence Ratio of LIGO's
time-shifted \emph{background} triggers, \emph{foreground} triggers and \emph{software injections}. This chapter describes how we have
adapted the methods from \citet{bcr_paper} in our implementation.

@@

Ideally, we would like to compute the BCRs for all triggers (foreground and background) produced by LIGO to help improve
the significance of astrophysically real triggers and decrease the significance of glitches. However, computational
limitations prevent analysis of all triggers. Instead, we limit the data to several 8-day chunks from O2, selected based
on the number of foreground triggers identified by \href{https://pycbc.org/}{\texttt{PyCBC}} in the various chunks, as seen in
Figure\textasciitilde\ref{fig:o2TrigCount}. Currently, we are using O2 data as it is public, whereas O3 data is not. By analyzing O2
triggers we can publish our results without any restrictions imposed by LIGO.

@@

\caption[BBH parameters corresponding to durations $<454$ ms]{\label{tab:parameters}Template Banks's parameters for templates with duration $<454$ ms.}
\centering
\begin{tabular}{lrr}
\toprule
  & Minimum & Maximmum\\
Component Mass 1 $[\text{M}_{\odot}]$ & 31.54 & 491.68\\
Component Mass 2 $[\text{M}_{\odot}]$ & 1.32 & 121.01\\
Total Mass $[\text{M}_{\odot}]$ & 56.93 & 496.72\\
Chirp Mass $[\text{M}_{\odot}]$ & 8.00 & 174.56\\
Mass Ratio & 0.01 & 0.98\\
\end{tabular}
\end{table}

@@

\caption[High-mass BCR search space.]{The template bank used by \href{https://pycbc.org/}{\texttt{PyCBC}} to search in O2's chunk 14. Our search is constrained to the high-mass parameter space enclosed by the dashed line (triggers with diurations \(<45s\).}\label{fig:templateBank}
\end{figure}

By restricting the \href{https://pycbc.org/}{\texttt{PyCBC}} triggers only based on the template duration, we acquire common (high FAR) and rare (low
FAR) triggers. It would be straightforward in principle to broaden our template-duration constraint to encompass all
triggers produced by all the various search pipelines. However, we refrain from doing so to keep our computational costs
manageable. A plot of the counts of the triggers in O2 after downsampling the triggers based on the template duration
can be viewed in Figure\textasciitilde\ref{fig:o2FilteredTrigCount}.

@@

We restrict the priors for Bayesian parameter estimation on the masses such that we only consider signals that are less
than 454ms in duration, as presented in Table\textasciitilde\ref{tab:parameters}. The spins are set to be aligned with a range from
{[}0,1{]}. The prior on luminosity distance assigns probability uniformly in comoving volume, with an upper cutoff of 5 Gpc.
The \href{https://pycbc.org/}{\texttt{PyCBC}} search that originally produced our set of triggers considered a more comprehensive range of masses and
spins than we do in the BCR computation for this project. To accommodate this, we prescreened the triggers to only allow
triggers with masses within our priors. The run time on a single trigger using the \href{https://arxiv.org/abs/1904.02180}{\texttt{dynesty}} implementation of nested
sampling, with \(2,000\) live points and \(100\) walkers is usually between \(0.5-12\) hours (where the speed often depends on
the SNR of the data segment).

@@

\begin{equation}
\mathrm{BCR} = \frac{\alpha Z^{S}}{\prod_{i=1}^{D}\left[\beta Z_{i}^{G}+(1-\beta) Z_{i}^{N}\right]}\ , \label{eq:bcr}
\end{equation}
where \(Z^S\) is the evidence for \(\mathcal{H}_S\), and \(Z^{G}_i\) and
\(Z^N_i\) are the evidences for \(\mathcal{H}_{Gi}\) and \(\mathcal{H}_{Ni}\),
in the \(i^{\text{th}}\) detectors. The arbitrary weights \(\alpha\) and \(\beta\)
parametrize our prior belief in each model:

\begin{itemize}
\item
  \(\alpha = P(\mathcal{H}_S)/P(\mathcal{H}_I)\)
\item
  \(\beta = P(\mathcal{H}_{Gi}|\mathcal{H}_I)=1-P(\mathcal{H}_{N}|\mathcal{H}_I)\)
\end{itemize}

for all \(i\) detectors. As most of the data should be noise, we expect both \(\alpha\) and \(\beta\), to be less than 1.
These parameters will be tuned to minimize the overlap between the signal and noise trigger populations.

@@

%%%%%%---ACKNOWLEDGEMENTS---%%%%%%%%%%%%
\begin{acknowledgments}
Helpful comments, OzGrav, LIGO, NSF

>>>>

\documentclass[%
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}

@@

@@

\begin{abstract}
The detection of high mass black holes ( $>100$ M${}_\odot$) will shed light on the formation of super-massive black holes and thus galaxy formation. Although LIGO is theoretically sensitive to the merger of binary black holes with total masses up to 500 M${}_\odot$, the largest total mass detected so far is approximately 80 M${}_\odot$. A possible explanation for the absence of high mass events may be due to their misclassification as short-duration instrumental noise transients. Short-duration instrumental transients mimic the short-duration gravitational-wave signals from high-mass binary black hole mergers. Here we demonstrate that a new search method utilising Bayesian inference could be a more sensitive tool for detecting high-mass binary black hole mergers as compared to traditional match-filtering. We have applied this technique on the high-mass triggers during LIGO's second observing run to investigate the possibility of discovering new gravitational-wave signals from an entirely new class of high mass black hole binaries.

@@

\begin{itemize}
    \item blip glitches and other short duration glitches occur very often
    \item searches have to down-weight short duration foreground triggers because of high occurrence rate of short duration glitches
    \item searches do not use CBC templates -- BCR can improve significance~\cite{BCR1}.
    \item BCR can help give true significance, or $p_{astro}$ of foreground triggers
\end{itemize}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{example-image-a}
    \caption{Some caption.}
    \label{fig:somefig}
\end{figure}

Since the 1970s, there has been an accumulation of evidence for stellar-mass and super-massive black holes. In 2019, the Event Horizon Telescope provided the first visual evidence of the super-massive black hole M87 \cite{event_horizon}.  As of January 2020, LIGO, Virgo and IAS have confirmed more than a dozen binary black hole systems \cite{GWTC1, IAS_2}. These discoveries have established the existence of stellar-mass black holes, super-massive black holes and binary black hole systems.  Interestingly, there is still no direct evidence for intermediate-mass black holes, the black holes that lie in between stellar-mass and super-massive black hole systems with masses between $10^2-10^6 \text{M}_\odot$. Thus, the existence of intermediate-mass black holes is still speculative.

If intermediate-mass black holes are present, the gravitational waves emitted from the merger of a binary black hole system with at least one intermediate-mass black hole (up to a mass of $400 \text{M}_\odot$) will be detectable by the ground-based gravitational-wave network. According to \citet{fregeau2006imbhbRatePrediction, mandel2008rates,rodriguez2015bbhRatePredictions}, gravitational waves from such systems within the detection of the ground-based detectors should occur at rates of one to up to ten events per year.  Hence it is peculiar that LIGO has not detected a gravitational wave event from a system with a total mass greater than $80 \text{M}_\odot$.

@@

There are two primary motivations for a targeted high-mass search that utilises Bayesian inference. The first is to help
distinguish gravitational-waves from high-mass systems from noise-transients, and the second is to try to alleviate the
discord in the significance of some of the ``quiet'' gravitational wave events between various search pipelines.

@@

\caption[Comparing glitches to high-mass candidates.]{Spectogram of a glitch on the left, and a candidate GW trigger on the right.}\label{fig:compareGlitchAndSignal}
\end{figure}

@@

The goal of this PhD's first project is to enhance the significance of high-mass foreground triggers using a Bayesian
approach and thereby increase the confidence in the detection of gravitational waves from large-mass compact binaries.
If this work does improve the significance of high-mass triggers, we will increase the likelihood of detecting a
high-mass black hole system via gravitational waves. The detection statistic we will utilise is the Bayesian Coherence
Ratio (or BCR), as seen in \citet{bcr_paper}. The BCR is discussed in detail in Appendix\textasciitilde\ref{BayesianCoherenceRatio}. In
short, The \texttt{BCR} is an odds (evidence ratio) comparing two hypotheses: there's a coherent GW in the data, plus Gaussian
noise vs there's only incoherent Gaussian noise or a glitch (modelled as a compact binary coalescence) in each detector
\citep{bcr_paper}.

@@

\caption[p-astro for various O2 foreground triggers]{\label{tab:O2significancesWObcr}p-astro from several detection pipelines for a subset of the O2 foreground triggers.}
\centering
\begin{tabular}{llrrrr}
\toprule
Event & Catalogue & PyCBC & GstLAL & cWB & IAS\\
\midrule
GW170104 & GWTC-1 & 1.00 & 1.00 & 1.00 & 0.99\\
GW170121 & IAS-1 & NA & NA & NA & 0.99\\
GWC170402 & IAS-2 & NA & 0.09 & NA & 0.68\\
GW170403 & IAS-1 & NA & NA & NA & 0.56\\
IMBHC170423 & IMBH-marginal & NA & 0.00 & 1.00 & NA\\
\addlinespace
GW170425 & IAS-1 & NA & NA & NA & 0.77\\
GW170729 & GWTC-1 & 0.52 & 0.98 & 0.94 & NA\\
\bottomrule
\end{tabular}
\end{table}

@@

\hypertarget{methodology}{%
\chapter{Methodology}\label{methodology}

This PhD project intends to provide new methods and insights to studying gravitational waves form high-mass systems.
Thus far, the project has focused on building up the infrastructure to compare the Bayesian Coherence Ratio of LIGO's
time-shifted \emph{background} triggers, \emph{foreground} triggers and \emph{software injections}. This chapter describes how we have
adapted the methods from \citet{bcr_paper} in our implementation.

@@

Ideally, we would like to compute the BCRs for all triggers (foreground and background) produced by LIGO to help improve
the significance of astrophysically real triggers and decrease the significance of glitches. However, computational
limitations prevent analysis of all triggers. Instead, we limit the data to several 8-day chunks from O2, selected based
on the number of foreground triggers identified by \href{https://pycbc.org/}{\texttt{PyCBC}} in the various chunks, as seen in
Figure\textasciitilde\ref{fig:o2TrigCount}. Currently, we are using O2 data as it is public, whereas O3 data is not. By analysing O2
triggers we can publish our results without any restrictions imposed by LIGO.

@@

\caption[BBH parameters corresponding to durations $<454$ ms]{\label{tab:parameters}Template Banks's parameters for templates with duration $<454$ ms.}
\centering
\begin{tabular}{lrr}
\toprule
  & Minimum & Maximum\\
\midrule
Component Mass 1 $[\text{M}_{\odot}]$ & 31.54 & 491.68\\
Component Mass 2 $[\text{M}_{\odot}]$ & 1.32 & 121.01\\
Total Mass $[\text{M}_{\odot}]$ & 56.93 & 496.72\\
Chirp Mass $[\text{M}_{\odot}]$ & 8.00 & 174.56\\
Mass Ratio & 0.01 & 0.98\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption[High-mass BCR search space.]{The template bank used by \href{https://pycbc.org/}{\texttt{PyCBC}} to search in O2's chunk 14. Our search is constrained to the high-mass parameter space enclosed by the dashed line (triggers with duration \(<45s\).}\label{fig:templateBank}
\end{figure}

By restricting the \href{https://pycbc.org/}{\texttt{PyCBC}} triggers only based on the template duration, we acquire common (high FAR) and rare (low
FAR) triggers. It would be straightforward in principle to broaden our template-duration constraint to encompass all
triggers produced by all the various search pipelines. However, we refrain from doing so to keep our computational costs
manageable. A plot of the counts of the triggers in O2 after down-sampling the triggers based on the template duration
can be viewed in Figure\textasciitilde\ref{fig:o2FilteredTrigCount}.

@@

We restrict the priors for Bayesian parameter estimation on the masses such that we only consider signals that are less
than 454ms in duration, as presented in Table\textasciitilde\ref{tab:parameters}. The spins are set to be aligned with a range from
{[}0,1{]}. The prior on luminosity distance assigns probability uniformly in comoving volume, with an upper cutoff of 5 Gpc.
The \href{https://pycbc.org/}{\texttt{PyCBC}} search that originally produced our set of triggers considered a more comprehensive range of masses and
spins than we do in the BCR computation for this project. To accommodate this, we screened the triggers to only allow
triggers with masses within our priors. The run time on a single trigger using the \href{https://arxiv.org/abs/1904.02180}{\texttt{dynesty}} implementation of nested
sampling, with \(2,000\) live points and \(100\) walkers is usually between \(0.5-12\) hours (where the speed often depends on
the SNR of the data segment).

@@

\begin{equation}
\mathrm{BCR} = \frac{\alpha Z^{S}}{\prod_{i=1}^{D}\left[\beta Z_{i}^{G}+(1-\beta) Z_{i}^{N}\right]}\ , \label{eq:bcr}
\end{equation}
where \(Z^S\) is the evidence for \(\mathcal{H}_S\), and \(Z^{G}_i\) and
\(Z^N_i\) are the evidences for \(\mathcal{H}_{Gi}\) and \(\mathcal{H}_{Ni}\),
in the \(i^{\text{th}}\) detectors. The arbitrary weights \(\alpha\) and \(\beta\)
parametrise our prior belief in each model:

\begin{itemize}
\tightlist
\item
  \(\alpha = P(\mathcal{H}_S)/P(\mathcal{H}_I)\)
\item
  \(\beta = P(\mathcal{H}_{Gi}|\mathcal{H}_I)=1-P(\mathcal{H}_{N}|\mathcal{H}_I)\)
\end{itemize}

for all \(i\) detectors. As most of the data should be noise, we expect both \(\alpha\) and \(\beta\), to be less than 1.
These parameters will be tuned to minimise the overlap between the signal and noise trigger populations.

@@

%%%%%%---ACKNOWLEDGEMENTS---%%%%%%%%%%%%
\begin{acknowledgments}
helpful comments, OzGrav, LIGO, NSF
